Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-03-29 doc/invt.tex
invt.tex: updates theory

\newcommand{\Err}{E}

@@

\begin{abstract}
  The Wang-Landau (WL) algorithm and metadynamics
  are two closely-related techniques
  in free energy calculations.
  %
  Both methods sample a flat distribution
  along a quantity of interest, $z$,
  by constructing a bias potential that offsets
  the potential of mean force (PMF) on the fly.
  %
  The two, however, differ by
  the manner of updating the bias potential:
  %
  the WL algorithm updates only
  the bin containing the current $z$,
  while metadynamics updates also
  several neighboring bins according to
  a Gaussian wave packet.
  %
  The error of the resulting PMF
  is determined by the updating schedule,
  or the updating magnitude
  as a function of simulation time.
  %
  In the WL case,
  the optimal schedule
  is known to be given by
  the inverse simulation time.
  %
  In this study,
  we give a formula of computing the optimal schedule
  for a general updating scheme
  involving multiple bins.
  %
  We show that while metadynamics can be advantageous
  for relatively short simulations,
  the WL algorithm is best for asymptotic convergence.
  %the optimal schedule
  %and the effectiveness of metadynamics
  %are sensitive to the simulation length and other parameters.
  %
  %Further, we compared the two algorithms, and found that
  %although metadynamics produces a smoother PMF profile,
  %it is less effective in reducing errors with the shorter wavelengths.
  %%
  %Thus, the latter relies more heavily on the assumption of
  %the smoothness of the PMF profile,
  %especially with a wide updating window.
\end{abstract}

@@

Free energy calculation\cite{frenkel, newman} is a central theme
in computational physics and chemistry.
%
Given a system,
the problem is to compute,
via either Monte Carlo\cite{frenkel, newman, landau_binder} (MC)
and molecular dynamics\cite{frenkel, karplus2002} (MD) simulations,
a distribution, $p^*(z)$,
along a quantity of interest $z$.
%
The negative logarithm,
$-\log p^*(z)$,
defines a free energy,
or the potential of mean force (PMF).
%
Normally, the distribution $p^*(z)$ is
limited in one or a few narrow $z$ regions.
%
Therefore, to capture the global shape
of the PMF,
one has to run multiple independent simulations
under different umbrella or bias potentials
along $z$\cite{mezei1987, berg1992, lee1993},
and then piece together the entire curve.
%
This solution is clumsy in practice
and often fails to converge quickly
for a complex and glassy system,
because the system can be trapped
in a local free energy minimum
for a long time.

@@

The Wang-Landau (WL) algorithm\cite{wang2001, wang2001pre}
and meta\-dy\-nam\-ics\cite{huber1994, laio2002,
laio2008, barducci2011, sutto2012}
are two widely-used and closely-related\cite{micheletti2004}
techniques for this purpose,
in which a bias potential is actively built up on-the-fly
during simulation.
%
These techniques employ updating loops
that discourage future visits to configurations
with the current $z$ value
by elevating the bias potential here.
%
The main difference between the two
is that in the WL case,
each updating step is limited to the bin
containing the current $z$,
while metadynamics adopts an extended
Gaussian wave packet
covering several neighboring bins.
%
The latter is more often used in MD simulations
as it guarantees a smooth profile
of the bias potential
suitable for numerical differentiation
in deriving the bias force.
%
However, in both cases, the bias potential
is updated frequently (usually every few MC or MD steps),
which brings about a side effect
of disrupting the underlying
equilibrium sampling dynamics\cite{laio2002}.
%
Thus, one often has to reduce
the updating magnitude
over the course of simulation
to decrease the error.

The updating schedule,
or the manner of reducing
the updating magnitude\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014,
barducci2008, dickson2011, dama2014},
therefore,
determines the precision of the final bias potential,
hence that of the PMF.
%
For the WL algorithm, it is known
that the optimal magnitude of updating $\ln f$
is given by the product of the
inverse time\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008}
(with ``time'' being
the number of simulation steps)
and the number of bins.
%
The optimal schedule for the metadynamics case,
however, is less studied.

@@

In analogous to Eq. \eqref{eq:x2t_average},
we have, from Eq. \eqref{eq:y2_sum},
the error after a period of $T$,
%
\begin{equation}
  \Err
  =
  \Err_R + \Err_A,
  \notag
  %\label{eq:error_split}
\end{equation}
%
which is the sum of the residual error
%
\begin{equation}
  \Err_R
  =
  \sum_{k = 2}^n
    \left\langle y_k^2(0) \right\rangle \,
    e^{ - 2 \, \lambda_k  \, q(T) },
  \label{eq:error_res}
\end{equation}
%
from the decay of the initial error,
and the asymptotic error
\begin{equation}
  \Err_A
  =
  \int_0^T \int_0^T
  \sum_{k = 2}^n
    \dot u_k\bigl( q(t) \bigr) \,
    \dot u_k\bigl( q(t') \bigr) \,
    \kappa_k(t - t') \, dt \, dt',
  \label{eq:error_asym}
\end{equation}
%
where,
\begin{align*}
  u_k(q')
  &\equiv
  e^{\lambda_k \, [q' - q(T)]},
  \\
  \kappa_k(t - t')
  &\equiv
  \left\langle
    \eta_k(t) \, \eta_k(t')
  \right\rangle.
\end{align*}
%
\note{The extremal condition of $\Err_A$ is given by
%
\begin{align}
\sum_{k = 2}^n
\lambda_k \, u_k\bigl( q(t) \bigr)
\int_0^T
  \ddot u_k \bigl( q(t') \bigr) \,
  \kappa_k(t' - t) \, dt' = 0.
\label{eq:optimal_mbin}
\end{align}
%
If all $\lambda_k$, hence all $u_k$, are the same,
the above condition can be satisfied
by Eq. \eqref{eq:ddu_eq_0},
which is the single-bin case.
%
Unfortunately, the above expression
appears to be useless in a general setting.
}

@@

Before addressing the optimal $\alpha(t)$,
let us first compute the equilibrium value of $\Err$
under a fixed schedule $\alpha(t) \equiv \alpha_0$.
%
In this case, we drop $\Err_R$, and
the error at the end of a long period $T'$
is given by
%
\begin{align}
  \Err
  &=
  \Err_A
  =
  \int_0^{T'}
    \sum_{k = 2}^n
      \Gamma_k \, (\lambda_k \, \alpha_0)^2 \,
      e^{ 2 \, \lambda_k \, \alpha_0 \, (t - T') }
    \, dt
  \notag
  \\
  &
  \stackrel{ T' \to \infty }
  { =\joinrel=\joinrel=\joinrel= }
  %
  \sum_{ k = 2 }^n
    \frac 1 2 \, \Gamma_k \, \lambda_k \, \alpha_0.
  \label{eq:error_eql}
\end{align}
%
In terms of the components, we have
%
\begin{equation}
  \left\langle
    y_k^2
  \right\rangle
  =
  \frac 1 2 \, \Gamma_k \, \lambda_k \, \alpha_0,
  \label{eq:y2_eql}
\end{equation}
%

Below we shall assume that before starting
the schedule $\alpha(t)$ to be optimized,
the system has entered the equilibrium
under a fixed $\alpha_0$.
%
That is, for $t \le 0$,
$\alpha(t) = \alpha_0$.
%
We can then use Eq. \eqref{eq:y2_eql}
for the $\langle y_k^2(0) \rangle$'s in Eq. \eqref{eq:error_res}.
%
\note{Another possible approximation
  is to interpret $\Gamma_k$ as twice the autocorrelation time,
  and the kernel is exponential
  $$
  \kappa_k(t) = \exp\left( - \frac{2}{\Gamma_k} |t| \right),
  $$
  then a slightly more accurate value is given by\cite{vankampen}
  $$
  \left\langle
    y_k^2
  \right\rangle
  =
  \frac{      ( \lambda_k \, \alpha_0 )^2     }
       { \lambda_k \, \alpha_0 + 2 / \Gamma_k }.
  $$
  But it appears that the main error of Eq. \eqref{eq:y2_eql}
  is an underestimate instead of an overestimate.
  And a possible source of error might be from
  converting a discrete sum to a continuous integral.
}
%
In this case,
the residual error at the end of period $T$
is given by
%
\begin{equation}
  \Err_R
  =
  \sum_{k = 2}^n
      \frac 1 2 \, \alpha_0 \, \Gamma_k \, \lambda_k \,
      e^{  - 2 \, \lambda_k \, q(T) }
  .
  \label{eq:error_res1}
\end{equation}
%
It follows that the residue error
depends on the schedule $\alpha(t)$
only through endpoint value of
the integral $q(T)$.
%
Thus, if we variate the schedule under a fixed $q(T)$,
then this part can be ignored,
and the optimization can be done exclusively
on the asymptotic part, $\Err_A$.

@@

As shown above, with fixed endpoint value of $q(t)$,
the optimization of the error is equivalent to
that of the asymptotic error.
%
We now rewrite Eq. \eqref{eq:error_asym1} as
%
\begin{equation}
  \Err_A
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt,
  \notag
  %\label{eq:error_asym_action}
\end{equation}
%
where, we have defined the Lagrangian,
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \dot u_k^2\bigl[ q(t) \bigr]
  \notag
  \\
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \lambda_k^2 \, u_k^2 \bigl[ q(t) \bigr]
  \; \dot q^2( t ).
  \label{eq:error_asym_lagr}
\end{align}
%
Note that the Lagrangian
does not explicitly depend on time, $t$,
so the Hamiltonian,
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \partial \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L,
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
is conserved,
and coincides with the Lagrangian.
%
So we may write
%
\begin{equation}
  \mathcal L
  =
    \textstyle\sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(t) \bigr]
  \;
  \dot q^2(t)
  =
  c^2
  \quad
  (c > 0).
  \label{eq:Lagrangian_const}
\end{equation}
%
Integrating the square root of the left-hand side, we get
%
\begin{equation}
  \int_{ 0 }^{ q(t) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  =
  c \, t
  ,
  \label{eq:q_opt}
\end{equation}
%
where $c$ is determined by
the boundary condition at $t = T$.
%
Thus, we have an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.

Although Eq. \eqref{eq:q_opt} is hard to solve explicitly,
we can obtain an expression regarding $\alpha(t)$
by differentiating Eq. \eqref{eq:Lagrangian_const},
which yields
%
\begin{equation}
  \frac{ d   }
       { d t }
  \left[
    \frac{       1     }
         { \alpha( t ) }
  \right]
  =
  \frac{
    \sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^3
      \, u_k^2 \bigl[ q(t) \bigr]
  }
  {
    \sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(t) \bigr]
  }
  .
  \label{eq:dinvadt}
\end{equation}
%
The right-hand side is a weighted average
of the eigenvalues, $\lambda_k$'s,
and we shall refer to this as the instantaneous
eigenvalue, $\lambda(t)$.

For the single-bin scheme, $\lambda_k = 1$,
and the instantaneous eigenvalue is also unity.
Thus,
$$
\frac { d   }
      { d t }
\left[
  \frac{      1    }
	   { \alpha(t) }
\right]
=
1,
$$
which again leads to Eq. \eqref{eq:alpha_invt1}.

For a general multiple-bin scheme,
we consider two limiting cases.
%
First, in the beginning of a long simulation,
we have $t \ll T$ and $q(T) - q(t) \to \infty$,
the sums on the right-hand side of Eq. \eqref{eq:dinvadt}
is dominated by the term with the largest
$u_k = e^{ \lambda_k \, [q(t) - q(T)] }$,
or, equivalently, with the smallest nonzero
$\lambda_k$.
Then
\begin{equation}
  \frac{ d   }
       { d t }
  \left[
    \frac{       1     }
         { \alpha( t ) }
  \right]
  =
  \lambda_{ \min }
  ,
  %\notag
  \label{eq:dinvadt_limit1}
\end{equation}
%
and
%
\begin{equation}
  \alpha( t )
  =
  \frac{               1               }
       { \lambda_{ \min } ( t + t_1 )  }
  .
  \label{eq:alpha_limit1}
\end{equation}
%

Second, near the end of simulation,
we have $t \to T$, $u_k \approx 1$,
%
\begin{equation}
  \frac{ d   }
       { d t }
  \left[
    \frac{       1     }
         { \alpha( t ) }
  \right]
  =
  \frac{
    \sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^3
  }
  {
    \sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
  }
  =
  \bar \lambda,
  %\notag
  \label{eq:dinvadt_limit2}
\end{equation}
%
where,
$\bar \lambda$
is the average of $\lambda_k$ weighted
by $\Gamma_k \, \lambda_k^2$,
and it tends to be dominated
by the largest $\lambda_k$.
%
Thus, in this regime, we have
%
\begin{equation}
  \alpha( t )
  =
  \frac{             1             }
       { \bar \lambda ( t + t_2 )  }
  .
  \label{eq:alpha_limit2}
\end{equation}
%
Thus, we may characterize the optimal schedule as
%
\begin{equation}
  \alpha(t)
  =
  \frac{                  1                     }
       { \lambda(t) \, \bigl[ t + t_0(t) \bigr] }
,
\notag
\end{equation}
%
where the instantaneous eigenvalue, $\lambda(t)$,
increases gradually from $\lambda_{\min}$ to $\bar \lambda$,
and $t_0(t)$ is a slowly-varying function.

@@

To use Eq. \eqref{eq:q_opt} in a real simulation,
we need to know both $\lambda_k$'s and $\Gamma_k$'s.
%
While the updating scheme alone
determines the eigenmodes and $\lambda_k$'s,
the values of $\Gamma_k$'s depend also on the system
and the underlying sampling scheme (MC or MD).
%
Below is a technique of measuring the latter.

The idea is that we can measure $\Gamma_k$'s
via a constant schedule, $\alpha(t) = \alpha_0$,
simulation using Eq. \eqref{eq:y2_eql}.
%
A practical difficulty is that
we often do not know the intrinsic distribution, $p^*_i$,
beforehand, and thus cannot compute
the shifted bias potential, $v_i$,
from the original one, $V_i$
(cf. Sec. \ref{sec:FDS}).
%
To circumvent the problem, we shall use
the following modification:
%
\begin{equation}
\operatorname{var} Y_k
=
\frac{1}{2}
\alpha \, \Gamma_k \, \lambda_k,
\label{eq:varY}
\end{equation}
%
where
``$\operatorname{var}$''
means the variance from time averaging,
and
%
\begin{align}
Y_k
&=
\sum_{ i = 1 }^n
  \phi_{k i} \, (V_i - \bar V),
\label{eq:Y_def}
\\
\bar V
&=
\sum_{ j = 1 }^n p_j \, V_j.
\end{align}

@@

\subsubsection{\label{sec:invt_alpha}
Inverse-time schedules}

The optimal schedule predicted
by the direct solution of Eq. \eqref{eq:q_opt}
is complicated for a general
multiple-bin updating scheme.
%
Several previous studies\cite{marsili2006, barducci2008, dickson2011}
have been focused on analytically simpler schedules.
%
As an example of these approaches,
we consider here a variant of
the inverse-time formula, Eq. \eqref{eq:alpha_invt1}
%
\begin{equation}
\alpha(t) = \frac{1}{\lambda \, (t + t_0) },
\label{eq:alpha_invtlambda}
\end{equation}
%
with an optimized proportionality constant, $\lambda$.
%
We shall compute the error computed from
this variationally optimized schedule
and compare it with that from the optimal schedule.
%
We shall show that the optimal parameter, $\lambda$,
is sensitive to the simulation time, $t$,
which parallels the shift of
the instantaneous eigenvalue,
$\frac{d}{dt}\alpha^{-1}(t)$,
as discussed in Sec. \ref{sec:optschedule}.

>>>>

\newcommand{\Err}{E}
\newcommand{\ii}{\mathrm{i}}

@@

\begin{abstract}
  The Wang-Landau (WL) algorithm and metadynamics
  are two closely-related techniques
  in free energy calculations.
  %
  Both methods sample a flat distribution
  along a quantity of interest, $z$,
  by constructing a bias potential that offsets
  the potential of mean force (PMF) on the fly.
  %
  The two, however, differ by
  the manner of updating the bias potential:
  %
  the WL algorithm updates only
  the bin containing the current $z$,
  while metadynamics updates also
  several neighboring bins according to
  a Gaussian wave packet.
  %
  The error of the resulting PMF
  is determined by the updating schedule,
  or the updating magnitude
  as a function of simulation time.
  %
  In the WL case,
  the optimal schedule
  is known to be given by
  the inverse simulation time.
  %
  In this study,
  we give a formula of computing the optimal schedule
  for a general updating scheme
  involving multiple bins.
  %
  We show that while metadynamics can be advantageous
  for relatively short simulations,
  the single-bin updating scheme
  used in the WL algorithm
  belongs to a class of updating schemes
  optimal for asymptotic convergence.
  %the optimal schedule
  %and the effectiveness of metadynamics
  %are sensitive to the simulation length and other parameters.
  %
  %Further, we compared the two algorithms, and found that
  %although metadynamics produces a smoother PMF profile,
  %it is less effective in reducing errors with the shorter wavelengths.
  %%
  %Thus, the latter relies more heavily on the assumption of
  %the smoothness of the PMF profile,
  %especially with a wide updating window.
\end{abstract}

@@

Free energy calculation\cite{frenkel, newman} is a central theme
in computational physics and chemistry.
%
Given a system,
the problem is to compute,
via either Monte Carlo\cite{frenkel, newman, landau_binder} (MC)
and molecular dynamics\cite{frenkel, karplus2002} (MD) simulations,
a distribution, $p^*(z)$,
along a quantity of interest $z$.
%
The negative logarithm,
$-\log p^*(z)$,
defines a free energy,
or the potential of mean force (PMF).
%
Normally, the distribution $p^*(z)$ is
limited in one or a few narrow $z$ regions.
%
Therefore, to capture the global shape
of the PMF,
one has to run multiple independent simulations
under different umbrella or bias potentials
along $z$\cite{mezei1987, berg1992, lee1993},
and then piece together the entire curve.
%
This solution can be clumsy in practice
and it often fails to converge quickly
for a complex and glassy system,
because the system can be trapped
in a local free energy minimum
for a long time.

@@

The Wang-Landau (WL) algorithm\cite{wang2001, wang2001pre}
and meta\-dy\-nam\-ics\cite{huber1994, laio2002,
laio2008, barducci2011, sutto2012}
are two widely-used and closely-related\cite{micheletti2004}
techniques for this purpose,
in which a bias potential is actively built up on-the-fly
during simulation.
%
These techniques employ updating loops
that discourage future visits to configurations
with the current $z$ value
by elevating the bias potential there.
%
The main difference between the two
is that in the WL case,
each updating step is limited to the bin
containing the current $z$,
while metadynamics adopts an extended
Gaussian wave packet
covering several neighboring bins.
%
The latter is more often used in MD simulations
as it guarantees a smooth profile
of the bias potential
suitable for numerical differentiation
in deriving the bias force.
%
However, in both cases, the bias potential
is updated frequently (usually every few MC or MD steps),
which brings about a side effect
of disrupting the underlying
equilibrium sampling dynamics\cite{laio2002}.
%
Thus, one often has to reduce
the updating magnitude
over the course of simulation
to decrease the error.

The updating schedule,
or the manner of reducing
the updating magnitude\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014,
barducci2008, dickson2011, dama2014},
therefore,
determines the precision of the final bias potential,
hence that of the PMF.
%
For the WL algorithm, it is known
that the optimal updating magnitude, $\ln f$,
is given by the product of the
inverse time\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008}
and the number of bins
(with ``time'' being
the number of simulation steps).
%
The optimal schedule for the metadynamics case,
however, is less studied.

@@

In analogous to Eq. \eqref{eq:x2t_average},
we have, from Eq. \eqref{eq:y2_sum},
the error after a period of $T$,
%
\begin{equation}
  \Err
  =
  \Err_R + \Err_A,
  \label{eq:error_split}
\end{equation}
%
which is the sum of the residual error
%
\begin{equation}
  \Err_R
  =
  \sum_{k = 2}^n
    \left\langle y_k^2(0) \right\rangle \,
    e^{ - 2 \, \lambda_k  \, q(T) },
  \label{eq:error_res}
\end{equation}
%
from the decay of the initial error,
and the asymptotic error
\begin{equation}
  \Err_A
  =
  \int_0^T \int_0^T
  \sum_{k = 2}^n
    \dot u_k\bigl( q(t) \bigr) \,
    \dot u_k\bigl( q(t') \bigr) \,
    \kappa_k(t - t') \, dt \, dt',
  \label{eq:error_asym}
\end{equation}
%
where,
\begin{align*}
  u_k(q')
  &\equiv
  e^{\lambda_k \, [q' - q(T)]},
  \\
  \kappa_k(t - t')
  &\equiv
  \left\langle
    \eta_k(t) \, \eta_k(t')
  \right\rangle.
\end{align*}
%
\note{The extremal condition of $\Err_A$ is given by
%
\begin{align}
\sum_{k = 2}^n
\lambda_k \, u_k\bigl( q(t) \bigr)
\int_0^T
  \ddot u_k \bigl( q(t') \bigr) \,
  \kappa_k(t' - t) \, dt' = 0.
\label{eq:optimal_mbin}
\end{align}
%
If all $\lambda_k$, hence all $u_k$, are the same,
the above condition can be satisfied
by Eq. \eqref{eq:ddu_eq_0},
which is the single-bin case.
%
Unfortunately, the above expression
appears to be useless in a general setting.
}

@@

Before addressing the optimal $\alpha(t)$,
we shall first consider a constant schedule
$\alpha(t) \equiv \alpha_0$,
and compute the equilibrium error, $\Err$,
in this case.
%
After a long period $T'$,
the residual error becomes negligible, and
%
\begin{align}
  \Err
  &=
  \Err_A
  =
  \int_0^{T'}
    \sum_{k = 2}^n
      \Gamma_k \, (\lambda_k \, \alpha_0)^2 \,
      e^{ 2 \, \lambda_k \, \alpha_0 \, (t - T') }
    \, dt
  \notag
  \\
  &
  \stackrel{ T' \to \infty }
  { =\joinrel=\joinrel=\joinrel= }
  %
  \sum_{ k = 2 }^n
    \frac 1 2 \, \Gamma_k \, \lambda_k \, \alpha_0.
  \label{eq:error_eql}
\end{align}
%
In terms of the components, we have
%
\begin{equation}
  \left\langle
    y_k^2
  \right\rangle
  =
  \frac 1 2 \, \Gamma_k \, \lambda_k \, \alpha_0.
  \label{eq:y2_eql}
\end{equation}
%

We shall use this result to model the initial error
before starting the schedule $\alpha(t)$ to be optimized.
%
That is, for $t < 0$, we assume
$\alpha(t) = \alpha_0$, and
the system has entered the equilibrium
under a fixed $\alpha_0$ at $t = 0$.
%
We can then use Eq. \eqref{eq:y2_eql}
for the $\langle y_k^2(0) \rangle$'s in Eq. \eqref{eq:error_res}.
%
\note{Another possible approximation
  is to interpret $\Gamma_k$ as twice the autocorrelation time,
  and the kernel is exponential
  $$
  \kappa_k(t) = \exp\left( - \frac{2}{\Gamma_k} |t| \right),
  $$
  then a slightly more accurate value is given by\cite{vankampen}
  $$
  \left\langle
    y_k^2
  \right\rangle
  =
  \frac{      ( \lambda_k \, \alpha_0 )^2     }
       { \lambda_k \, \alpha_0 + 2 / \Gamma_k }.
  $$
  But it appears that the main error of Eq. \eqref{eq:y2_eql}
  is an underestimate instead of an overestimate.
  And a possible source of error might be from
  converting a discrete sum to a continuous integral.
}
%
Then,
the residual error at the end of period $T$
is given by
%
\begin{equation}
  \Err_R
  =
  \sum_{k = 2}^n
      \frac 1 2 \, \alpha_0 \, \Gamma_k \, \lambda_k \,
      e^{  - 2 \, \lambda_k \, q(T) }
  .
  \label{eq:error_res1}
\end{equation}
%
Note that the residual error
depends on the schedule $\alpha(t)$
only through endpoint value of
the integral $q(T) = \int_0^T \alpha(t) \, dt$.
%
Thus, if we variate the schedule under a fixed $q(T)$,
this residual part can be ignored,
and the optimization can be done exclusively
on the asymptotic part, $\Err_A$.

@@

As shown above, with fixed endpoint value, $q(T)$,
the optimization of the total error is equivalent to
that of the asymptotic part.
%
We can rewrite Eq. \eqref{eq:error_asym1} as an action
%
\begin{equation}
  \Err_A
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt,
  \notag
  %\label{eq:error_asym_action}
\end{equation}
%
where the Lagrangian,
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \dot u_k^2\bigl[ q(t) \bigr]
  \notag
  \\
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \lambda_k^2 \, u_k^2 \bigl[ q(t) \bigr]
  \; \dot q^2( t ),
  \label{eq:error_asym_lagr}
\end{align}
%
represents a free particle
with a position-dependent mass.
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian,
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \partial \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L,
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
which coincides with the Lagrangian,
is conserved.
%
So we may set
%
\begin{equation}
  \mathcal L
  =
    \textstyle\sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(t) \bigr]
  \;
  \dot q^2(t)
  =
  c^2
  \quad
  (c > 0).
  \label{eq:Lagrangian_const}
\end{equation}
%
Integrating the square root of the left-hand side, we get
%
\begin{equation}
  \int_{ 0 }^{ q(t) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  =
  c \, t
  ,
  \label{eq:q_opt}
\end{equation}
%
where $c$ is to be determined by
the boundary condition at $t = T$ as
%
\begin{equation}
  c =
  \frac 1 T
  \int_{ 0 }^{ q(T) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  .
  \notag
  %\label{eq:q_opt}
\end{equation}
%
Thus, we have an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.

Unfortunately,
Eq. \eqref{eq:q_opt} does not provide much information
on the behavior of $\alpha(t)$,
and the explicit solution
can be formidable in the general case.
%
We can nonetheless obtain an expression regarding $\alpha(t)$
by differentiating Eq. \eqref{eq:Lagrangian_const},
which yields
%
\begin{equation}
  \frac{ d   }
       { d t }
  \left[
    \frac{       1     }
         { \alpha( t ) }
  \right]
  =
  \frac{
    \sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^3
      \, u_k^2 \bigl[ q(t) \bigr]
  }
  {
    \sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(t) \bigr]
  }
  \equiv
  \lambda(t)
  .
  \label{eq:dinvadt}
\end{equation}
%
The expression in the middle is a weighted average
of the eigenvalues, $\lambda_k$'s,
and we shall refer to it as the instantaneous
eigenvalue, $\lambda(t)$.
%
The instantaneous eigenvalue
gives the rate of change of
the inverse updating magnitude.

For the single-bin scheme, $\lambda_k = 1$,
and the instantaneous eigenvalue is always unity.
Thus,
$$
\frac { d   }
      { d t }
\left[
  \frac{      1    }
       { \alpha(t) }
\right]
=
1,
$$
which again leads to Eq. \eqref{eq:alpha_invt1}.

For a general multiple-bin scheme,
we can characterize the solution of
Eq. \eqref{eq:dinvadt} as
\begin{equation}
  \alpha(t)
  =
  \frac{                  1                     }
       { \lambda(t) \, \bigl[ t + t_0(t) \bigr] }
,
  \label{eq:alpha_approxinvt}
\end{equation}
%
where
$t_0(t)$ is a slowly-varying function.

By differentiating Eq. \eqref{eq:dinvadt},
we can show that instantaneous eigenvalue,
$\lambda(t)$, is an increasing function.
%
\note{Proof that $\lambda(t)$ is an increasing function.
  Using $\dot u_k = \lambda_k \, u_k \, \alpha(t)$,
  we have
  $$
  \begin{aligned}
    \dot \lambda
    &=
    \frac
    {
      \sum_{k = 2}^n 2 \, \Gamma_k \, \lambda_k^3 \, u_k \, \dot u_k
    }
    {
      \sum_{k = 2}^n \Gamma_k \, \lambda_k^2 \, u_k^2
    }
    -
    \frac
    {
      \sum_{k = 2}^n \Gamma_k \, \lambda_k^3 \, u_k^2
      \sum_{k = 2}^n 2 \, \Gamma_k \, \lambda_k^2 \, u_k \, \dot u_k
    }
    {
      \left(
        \sum_{k = 2}^n \Gamma_k \, \lambda_k^2 \, u_k^2
      \right)^2
    }
    \\
    &=
    \frac
    {
      \sum_{k = 2}^n \Gamma_k \, \lambda_k^4 \, u_k^2
      \,
      \sum_{k = 2}^n \Gamma_k \, \lambda_k^2 \, u_k^2
      -
      \left(
        \sum_{k = 2}^n \Gamma_k \, \lambda_k^3 \, u_k^2
      \right)^2
    }
    {
      \left(
        \sum_{k = 2}^n \Gamma_k \, \lambda_k^2 \, u_k^2
      \right)^2
    }
    2 \alpha(t).
  \end{aligned}
  $$
  The numerator of the fraction of the last expression
  is nonnegative by the Cauchy-Schwarz inequality,
  or by the nonpositiveness of the discriminant
  of the following nonnegative definite quadratic form of $X$:
  $$
  \sum_{k = 2}^n \Gamma_k \, \lambda_k^2 \, u_k^2 \, (X - \lambda_k)^2 \ge 0.
  $$
  Thus, $\dot \lambda \ge 0$, and $\lambda(t)$ is monotonically increasing.
}
%
Below we shall study the initial and final values
of $\lambda(t)$.
%
At the end of a simulation,
we have $t = T$, and $u_k = 1$.
%
Thus,
the instantaneous eigenvalue
%
\begin{equation}
  \lambda(T)
  =
  \frac{
    \sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^3
  }
  {
    \sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
  }
  ,
  \label{eq:dinvadt_limit2}
\end{equation}
is an average of $\lambda_k$'s weighted
by $\Gamma_k \, \lambda_k^2$,
and it is independent of the simulation length, $T$.
%
Further, it often occurs that $\Gamma_k$
increases with $\lambda_k$,
such that the relative weight $\Gamma_k \, \lambda_k^2$
also increases with $\lambda_k$.
%
It follows that $\lambda(T)$ will be dominated
the larger values of $\lambda_k$'s.

By contrast,
at the beginning of a simulation,
the relative weight, $\Gamma_k \, \lambda_k^2 \, u_k^2(0)$,
is heavily influenced by the exponential factor
$u_k^2(0) = e^{ -2 \, \lambda_k \, q(T) }$,
and hence is sensitive to the simulation length, $T$.
%
A longer simulation means a smaller $\lambda(0)$,
and that $\lambda(t)$ would be small
for longer near the beginning of the simulation.

@@

To use Eq. \eqref{eq:q_opt} in a real simulation,
we need to know both $\lambda_k$'s and $\Gamma_k$'s.
%
While the updating scheme alone
determines the eigenmodes and $\lambda_k$'s,
the values of $\Gamma_k$'s depend also on the system
and the underlying sampling process (MC or MD).
%
Below is a technique of measuring $\Gamma_k$'s.

The idea is to measure $\Gamma_k$'s
via a simulation with a constant schedule,
$\alpha(t) = \alpha_0$,
using Eq. \eqref{eq:y2_eql}.
%
A practical difficulty is that
we often do not know the intrinsic distribution, $p^*_i$,
beforehand, and thus cannot compute
the shifted bias potential, $v_i$,
from the original one, $V_i$
(cf. Sec. \ref{sec:FDS}).
%
To circumvent the problem, we shall use
the following modification:
%
\begin{equation}
\operatorname{var} Y_k
=
\frac{1}{2}
\alpha \, \Gamma_k \, \lambda_k,
\label{eq:varY}
\end{equation}
%
where
``$\operatorname{var}$''
means the variance from time averaging,
and
%
\begin{align}
Y_k
&=
\sum_{ i = 1 }^n
  \phi_{k i} \, (V_i - \bar V),
\label{eq:Y_def}
\\
\bar V
&=
\sum_{ j = 1 }^n p_j \, V_j.
\notag
\end{align}

@@
